Searching for standards of fairness in the transportation justice literature: A scoping review

### 2023-11-15 Manuscript/Manuscript.qmd
Incorporated Matt Palm's executive summary (EC), revised and inserted comments in EC, made minor edits to body and to Glossary, and added acknowledgements section

---
title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
bibliography: All_papers_in_extracted.bib
format: docx
---

@@

```{r, eval=FALSE}
Pflow_chart %>%
  DiagrammeRsvg::export_svg() %>% 
  charToRaw() %>% rsvg::rsvg_png(file="Pflow_chart.png")
```

@@

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())
#+ ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
#  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("standards_conc_percentages_plot.png")
```

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
#standards_conc_percentages_plot_and_table
# ggsave("standards_conc_percentages_plot_and_table.png")
```

@@

**NOTE:** After brainstorming, Anastasia, Ignacio, and Antonio suggest that an executive summary could be a good strategy for readers who are too busy to read, and to provide a good idea of what readers who do plan to read will find in this document.

_If we plan to write an executive summary, it should ease readers into the report. The full report is long so this may be a useful 2-3 page summary addition._

@@

_This portion may be helpful to guide readers to desired sections, if they are pressed for time or interested in only particular parts of the report. For instance:_

_- If you're interested in why we wrote this report read the "Introduction."_  
_- Curious about the conceptual framework for transport justice? Read the framework in "A framework to analyze questions of justice"_  
_- Want to know about our calls for action to move transport equity analysis and standard-setting substantively towards questions of transport justice? Read the "Moving forward: concluding remarks and calls for action". This section is geared to both researchers and decision-makers._

@@

# Glossary
_This section may be helpful for readers to clarify our definitions of these terms so they can more easily follow the text and as a quick reference:_

- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values. (*maybe add this into the intro??*)
- **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
- **Justice**: mutually reinforced by the tools of fairness and equity, justice is a multi-dimensional political ideal attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. It ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440]. The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
- **The temporal-spatial context of justice**: the answers to "When?" and "Where?"
- **The subjects of justice**: the answer to "Who?" is the focus of justice.
- **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice. 
- **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
- **The rationale to justice**: the "Why?". Connects with the context, subject, object and methods used for justice determination.
- **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are s that when operationalized effectively define what is fair.
- **Accessibility**:  
- **Capabilities approach**:  
- **Equality**:  
- **Mobility**:  
- **Policy**:  
- **Social exclusion**:  
- **Space-time convergence**:  
- **Sufficientarianism**:  
- **Theories of justice**:  
- **Utilitarianism**:  
Etc.,

# Justice in transportation?

@@

# Setting the stage

@@

## Justice, equity, fairness, and standards

@@

## A framework to analyze questions of justice

@@

# Scanning the lay of the land: Methods

@@

1. An initial limited search of Web of Science (WoS) Core Collection was undertaken to identify key documents This collection contains documents in journals, conference proceedings, and books published all over the world. Separate sarches using the terms 'transportation' and 'equity' were generated. From these searches, we examined the text contained in the titles and abstracts of relevant papers, the index terms used to describe the papers, and subject heading searches when available. As we developed a clearer outline of the literature, we continued with this process by adjusting the terms used for the search. This took the form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Fair"...**2**), where **1** and **2** signify additional terms relating to 'transportation' and 'equity', respectively.

@@

1. The second step (green box in @fig-fig1) included scanning all full-text papers which passed step 1. These papers were reviewed to determine if they included a relevant "how", i.e., an standard and/or relevant theoretical or conceptual discussion. At this stage, papers were evaluated again by two independent reviewers who voted for inclusion or exclusion. If an article was voted to be excluded, it was tagged with one of five possible reasons for exclusion, namely (1) no standards included; (2) no relevant conceptual elements included; (3) no standard and no conceptual elements included; (4) send back -- QA issue; or (5) other. Discrepancies were resolved by a third reviewer.

1. In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-fig1). The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" (equity standard(s) and concepts) was filled out for each study. @fig-A2 contains the template that was input into *Covidence* and used throughout.

@@

# The lay of the land: a summary of findings

@@

```{r plotting-year-cont-plot-fig3}
#| label: fig-fig3
#| fig-cap: "Papers included in the review by year of publication and case study continent."
year_cont_plot
ggsave("year_cont_plot.png")
```

@@

```{r plotting-population-group-plot-fig4}
#| label: fig-fig4
#| fig-cap: "The proportion of papers that focus on each type of population group. Categories for population groups were generated upon data extraction."
pop_plot
ggsave("pop_plot.png")
```

@@

```{r plotting-mode-plot-fig5, fig.width=5}
#| label: fig-fig5
#| fig-cap: "The proportion of papers that investiage each type of mode. Categories for modes were generated upon data extraction."
mode_plot
ggsave("mode_plot.png")
```

@@

In terms of papers that study specific destinations, the activities most commonly studied include healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`), followed by employment (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`). Papers that exclusively focus on healthcare typically originate from the healthcare planning literature, and look to inform planners about disparities in services and corrective actions. For instance, @wangModelingSpatialAccessibility2011 model access of Chinese-language speakers in Toronto (Canada) to Mandarin-speaking family physicians. These authors infer an inequity from the spatial mismatch between language competent healthcare providers and Chinese speakers. Papers that exclusively focus on employment typically focus on these trips as they are the most common trip purpose and are often correlated with other trip activities like shops, recreation, and other services. For instance, @allenSizingTransportPoverty2019a propose a low employment-based accessibility threshold and a composite population vulnerability index to identify transportation-poor neighborhoods in urban Canada. Papers that focus on healthcare and employment typically source data from representative travel surveys/diaries, census data, and points-of-interest databases. In other words, these studies often benefit from well developed and institutional data that represents 'typical trips', especially in the Global North where these data are more readily available.

@@

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig6
#| fig-cap: "The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category."
standards_conc_percentages_plot_and_table
ggsave("standards_conc_percentages_plot_and_table.png")
```
<!--`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards? what types of methods?", ".")`-->
We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice. Broadly speaking, some trends emerge in terms of the methods used depending on the way equity is conceptualized and the type of standard suggested; See @fig-fig6 for definitions of of conceptualizations along the horizontal axis and types of standards along the vertical axis. These categories emerged as we reviewed the literature and we use them to discuss trends; they are by no means an exhaustive list of equity conceptualizations and types of standards.

@@

# A critical appraisal of the lay of the land

@@

Most papers within our corpus (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North. Though their subject matter is varied, their spatial context mainly pertains to North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; or autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).

@@

Temporally, we see the literature from the Global North and Global South in our corupus as laying on different transport inequity continuums. Literature from the Global North is concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits e.g., [@cuiSpatialAccessPublic2020]. Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures NO2 road-side air pollution and find no levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups. These international guidelines, while perhaps being relevant at a time of lower transport-related development, are no longer relevant for some areas in the Global North. Along this line, inequities in the Global North are rising - high-income countries contain the wealthiest populations that have the most mobility (and produce associated carbon externalities) [@chancelCarbonInequalityKyoto2015].

@@

So what's missing from the "who?" question? We believe, more nuisances perspectives that capture the "who" of inequities is needed. We must go beyond the focus on the low-income transit rider; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.

@@

# Moving forward: calls for action

@@

# Final remarks
<!-- HERE IS AN OPPORTUNITY HERE TO CIRCLE BACK TO BEZOS SUBORBITAL FLIGHT, TO CLOSE THE REPORT WITH A REMINDER OF WHY IT IS IMPORTANT TO LEARN ABOUT JUSTICE.-->

We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people. We wondered why coverage of the event, in particular Bezos declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to? All of the human species or only those who can afford space tourism? What are the benefits of space tourism and how are they distributed? Who are the recipients of the impacts?

@@

@@

# Appendix

>>>>

---
title: "Searching for standards of fairness in the transportation justice literature: A scoping review"
subtitle: "Searching for standards of fairness"
author:
  - name:
      given: Anastasia
      family: Soukhov
    affiliations:
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: soukhoa@mcmaster.ca
  - name:
      given: Ignacio
      family: Tiznado-Aitken
    affiliations:
      - name: University of Toronto Scarborough
        department: Department of Geography and Planning
    email: i.tiznadoaitken@utoronto.ca
  - name:
      given: Matthew
      family: Palm
    affiliations:
      - name: University of North Carolina Chapel Hill
        department: Department of City and Regional Planning
    email: palmmatt@unc.edu
  - name:
      given: Steven
      family: Farber
    affiliations:
      - name: University of Toronto Scarborough
        department: Department of Geography and Planning
    email: steven.farber@utoronto.ca
  - name:
      given: Antonio
      family: Paez
    affiliations:
      - name: McMaster University
        department: School of Earth, Environment and Society
    email: paezha@mcmaster.ca
bibliography: All_papers_in_extracted.bib
format: docx
---

@@

```{r, eval=FALSE}
Pflow_chart %>%
  DiagrammeRsvg::export_svg() %>% 
  charToRaw() %>% rsvg::rsvg_png(file="figures/Pflow_chart.png")
```

@@

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())
#+ ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
#  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("figures/standards_conc_percentages_plot.png")
```

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
#standards_conc_percentages_plot_and_table
# ggsave("figures/standards_conc_percentages_plot_and_table.png")
```

@@

Modern societies provide people with varying degrees of mobility and accessibility, from low wage Amazon workers’ depending on infrequent bus service, to Jeff Bezos taking private flights into space. Societies implicitly tolerate different levels of inequality in access, though these standards are occasionally made legible through policy standards. For example, U.S. transit agencies are required to set and adhere to maximum levels of disparity in service quality between minority and non-minority routes as defined by the U.S. Civil Rights Act. In contrast to prior reviews on how to measure inequality in transportation systems, this report is concerned with the implied or explicit standards that are used to judge whether the measured inequalities are fundamentally “fair” or unacceptable. The present study achieves this objective by scanning the state of academic knowledge in defining “fairness” in the transport domain.

Borrowing from philosophers of justice [@jaggarPhilosophicalChallengesGlobal2009], the report’s authors ask the following questions about existing standards of fairness in practice:

-	**Where and when** are equity standards applied within the research depending on location and time of publication? How are context-specific planning processes changing?
-	**Who** benefits and is burdened by transportation systems? Along which lines of identity are transportation inequities examined?
-	**What** are those benefits and burdens of transportation systems? Are they qualified in sidewalk widths, bus frequency, travel times, accessibility, or some other consideration?
-	**How**: are equity standards and inequities measured? Are they infrastructure provision thresholds, acceptable pollution levels, or some other kind of cut off? How are these studies born out of broader notions of justice, i.e. disability rights, environmental justice, egalitarianism, or something else?

**Methods**
The authors conduct a scoping review of the literature. The review involves entering combinations of equity-related keywords into academic databases to identify relevant studies. The team initially identified 6,832 references from three diverse databases that they shrunk to a final sample using a four-step process:

1.	Removing 611 duplicate studies.
2.	Having a team of trained students review the remaining titles and abstracts for relevance, with 4,511 studies excluded for not sufficiently addressing transportation equity. 
3.	Reviewing the full-texts of the remaining 1,711 studies. At least two trained students reviewed each document, with authors breaking any ties. This resulted in 1,545 additional studies being excluded. At this stage, studies could be excluded if they did not include any standard for, or conceptualization of, equity.
4.	Extracting data on the "When", "Where", "Who", "What", and "How" of transport equity in each study. This included noting methods used, standards suggested, and inferring conceptualizations of equity that may connect to broader notions of justice from the remaining 165 studies.

**Results**
**When and Where: the contexts of justice**. Most (60%) studies deploying standards for equity are published on the Global North, including cases from Europe, the U.S., Canada, Australia, New Zealand, Japan, and Israel.  These studies focus on groups seen as disadvantaged within national contexts, with standards related to national policies or issues. In contrast, research on the Global South tends to adopt international standards in analyses, such as the U.N.’s Sustainable Development Goals (SDGs). Within the Global South, studies most commonly cover Asia, followed by Latin America. Studies in the latter region focus disproportionately on mobility barriers and the financial affordability of transportation. Studies pertaining to Africa are numbered, though they cover a broader scope of issues including informal transportation, citizens’ accessibility needs, and infrastructure development needs. 
Across the globe, the reviewed studies apply equity standards overwhelmingly to urban contexts (85%) and are concerned with a focused set of issues around. Studies on rural contexts are broader in the topics examined, from the equity of ferry service in the Philippines to the impacts of road construction on remote Amazonian communities.

**Who: the subjects of justice.** Income groups are studied most, followed by age groups, and people with disabilities. People with low incomes generally experience the least benefits and greatest burdens of transportation systems, including regarding costs and pollution exposure. Studies examining equity with respect to age demonstrate how older adults and children face different levels of access to key destinations due to differences in their capabilities. Many studies apply universal design guidelines and laws as equity standards to evaluate the equity of transportation systems for people with disabilities. Finally, researchers also frequently develop composite indices of vulnerability to identify neighbourhoods with high concentrations of multiple vulnerable populations. This approach is common when examining environmental justice impacts.

<!--We suggest: Who and What: the tools of mobility. I like our more-->
**What: the tools of mobility.** Most studies deploying equity standards apply them to either transit (37%) or pedestrian modes (25%). Many transit analyses are multimodal in nature, comparing the access transit provides against the access automobiles provided as an implicit standard for comparison. Others combine transit with non-auto modes to examine access for households without vehicles. These studies focus on delivering equitable mobility, sometimes in context of also meeting climate goals. However, they often fail to include explicit minimums of service that should be provided to different populations. In contrast, rights-based, infrastructure-focused studies apply principles of universal design in the study of transport systems and related infrastructure.

<!--we suggest: What: the benefits of mobility and accessibility. I like ours more; there is a distinction. -->
**What: the benefits of mobility.** Studies measure either movement (i.e., trips taken, trip quality, etc.) or the potential for movement (i.e. accessibility to specific destinations, or generally), with most studies focused on the latter. Over a quarter of the accessibility focused papers do not focus on access to specific destinations, but instead on the accessibility of the trip (i.e., level of service, barriers of infrastructure) or bundles of destinations. Studies that center particular destinations that should be reached/accessed include: employment, healthcare, grocery stores, and greenspace. Access to community support or places of worship, as well as childcare activities are less common in the reviewed literature.

<!--we suggest: "How": concepts of justice and standards of fairness. I like ours more - there is a distinction. -->
**How: concepts and standards of justice.** The most common standards are defined using access to opportunities cut offs (37.2%), opportunity standards, or thresholds related to the perceptions or activities of specific populations (35.8%), population standards. The latter often includes standards informed by public health, such as the number of physical activities undertaken in a week and perceived life satisfaction. Opportunity standards are evaluated through horizontal equity and spatial equity lenses, while population measures are evaluated through well-being frameworks. Another category that emerged in the literature are infrastructure-based standards, which include adherence to universal design guidelines or specific benchmarks for road network characteristics; they are less common in the literature comparatively and are often framed in rights-based terms, particularly as they relate to achieving universal design. But infrastructure standards around sustainable infrastructure (i.e. block lengths or distance between transit stops) have also been evaluated through vertical, horizontal and spatial lenses. The distribution of studies with respect to the kinds of standards deployed, and the broader concepts of justice used, are presented in the Figure below:

![The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category.](figures/standards_conc_percentages_plot_and_table.png)

**Calls to Action**

From what does and does not exist in the state of knowledge today, the authors challenge practitioners, researchers, advocates, and government to commit to five calls to action:

<!--we suggest:
Call 1: The need for explicit conceptualizations of equity to ground standards. I like Matts!-->
Call 1: Underpin standards in rigorous concepts of justice.

The standards of equity used in research are often weakly linked to any legally, intellectually, or publicly accepted conceptualizations of justice. Many standards are arbitrary. For example, the 15-minute city sets a threshold that is easy to remember, but is this really what is needed to achieve sufficient sustainable mobility for urban residents, as opposed to a 10 or 20 minute city?  And what conceptualization of human need should determine what services are essential enough to need to be within this threshold, and for whom?

<!--we suggest:
Call 2: The need for creative systems-thinking approaches to measuring (in)equities I like Matts!-->
Call 2: Develop creative methods for systems-thinking approaches to equity.

Mixed-methods research can help government set meaningful standards by better linking quantitative thresholds with qualitative assessments of well-being, freedom of movement and perceived access, among other dimensions. Yet most of the reviewed research is purely quantitative or, to a much lesser degree, purely qualitative. The constant deployment of metrics Lorenz Curves of Gini coefficients without research on where such measures need to be to improve quality of life (i.e., a standard), will not help governments with limited resources achieve the best possible equity outcomes within their respective contexts. Mobilizing Justice’s Priority Populations working group was inaugurated to advance new research in this area.

<!-- We suggested:
Call 3: Data availability as a matter of justice-->
Call 3: Data availability as a matter of justice.

Several areas of critical importance, such as access to care and community, are under-researched, and the authors suspect this relates to lack of data on these kinds of destinations. Furthermore, data issues prevent the application of many methods of equity analyses in the Global South. The production, availability, transparency, and ownership of data is a matter of justice on which governments and NGOs should act. In the Canadian context, Mobilizing Justice’s National Survey working group will advance data practices by conducting the first national survey on transportation poverty.

<!-- We suggested:
Call 4: Direct and explicit links between standards and experienced outcomes. Sure I like Matt's more!-->
Call 4: Develop more direct and explicit links between standards and lived experiences.

Linking standards to lived experiences and individual outcomes can help increase their likelihood of adoption by governments. Further, they can help decision makers understand the role of mobility in improving policy outcomes in other domains like nutrition, physical and mental health, education, and human development.  These links can help make the standards used in research and policy more comprehensible and relatable to community members and people with lived experience navigating unequal mobility systems. Mobilizing Justice launched it’s Data Driven Standards working group to advance research in this area.

<!-- We suggested:
Towards evaluations of interventions and policies in and across communities-->
Call 5: Rigorously evaluate interventions and policies.

Approximately 10% of the reviewed studies assessed interventions. Delivering new equity-focused policies and programs that do not actually advance equity harms communities, metastasizes mistrust, and delegitimates future efforts to promote transportation justice. Furthermore, policies that benefit one group of people in one context may not benefit other groups in other situations. A deep breadth of equity evaluations is needed to avoid failure and guide policymakers towards interventions that will work for the communities they serve. Mobilizing Justice launched its Innovative Pilots working group to begin filling this gap, through further work is needed.

@@

If you're interested in the motivation behind this report, read the first section, [Justice in transportation?](#sect1)

Keen to jump straight to a description of the conceptual framework for transport justice? See the second section, particularly the subsection: [A framework to analyze questions of justice](#sect2.2).

Interested in the methodological details of the literature review? See [section three](#sect3) and extended information in the [Appendix](#sect8)

Care more about the results and the critical appraisal of the results from the review? See sections [four](#sect4) and [five](#sect5), respectively.

Want to know about our calls for action to move transport equity analysis and standard-setting substantively towards questions of transport justice? Read the sixth section [Moving forward: calls for action](#sect6) and the final concluding section [Final remarks](#sect7). These sections are geared to both researchers and decision-makers.

# Glossary

- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values.
- **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
- **Justice**: mutually reinforced by the tools of fairness and equity, justice is a multi-dimensional political ideal attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. It ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440]. The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
- **The temporal-spatial context of justice**: the answers to "Where?" and "When?". What regions, within what countries, spatial contexts, point of time, and processes of planning are impacting transport systems?
- **The subjects of justice**: the answer to "Who?" is the focus of justice.
- **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice. 
- **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
- **The rationale to justice**: the "Why?". The rationale connects with the context, subject, object and methods used for justice determination.
- **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are something that when operationalized effectively define what is fair.
- **Accessibility**:  
- **Capabilities approach**:  
- **Equality**:  
- **Mobility**:  
- **Policy**:  
- **Social exclusion**:  
- **Space-time convergence**:  
- **Sufficientarianism**:  
- **Theories of justice**:  
- **Utilitarianism**:  
Etc.,
<!--NOTE: not too sure how many of these terms we need.. or if we need a glossary at all. Perhaps we can ask peer-reviewers about this need.-->

@@

# Justice in transportation? {#sect1}

@@

# Setting the stage {#sect2}

@@

## Justice, equity, fairness, and standards {#sect2.1}

@@

## A framework to analyze questions of justice {#sect2.2}

@@

# Scanning the lay of the land: Methods {#sect3}

@@

1. An initial limited search of Web of Science (WoS) Core Collection was undertaken to identify key documents This collection contains documents in journals, conference proceedings, and books published all over the world. Separate searches using the terms 'transportation' and 'equity' were generated. From these searches, we examined the text contained in the titles and abstracts of relevant papers, the index terms used to describe the papers, and subject heading searches when available. As we developed a clearer outline of the literature, we continued with this process by adjusting the terms used for the search. This took the form: ("Transport" OR "Transit" OR "Car\*" OR "Walk" OR "Bike"...**1**) AND ("Equity" OR "Justice" OR "Fair"...**2**), where **1** and **2** signify additional terms relating to 'transportation' and 'equity', respectively.

@@

2. The second step (green box in @fig-fig1) included scanning all full-text papers which passed step 1. These papers were reviewed to determine if they included a relevant "how", i.e., an standard and/or relevant theoretical or conceptual discussion. At this stage, papers were evaluated again by two independent reviewers who voted for inclusion or exclusion. If an article was voted to be excluded, it was tagged with one of five possible reasons for exclusion, namely (1) no standards included; (2) no relevant conceptual elements included; (3) no standard and no conceptual elements included; (4) send back -- QA issue; or (5) other. Discrepancies were resolved by a third reviewer.

3. In the last step, a data extraction template for each record was filled by one reviewer (purple box in @fig-fig1). The data extraction template was created with the aim of striking a balance between the complexity of categories and the simplicity of summary; information related to "Where?" (the geographical context and sphere of life), "When?" (temporal circumstances for the application of justice), "Who?" (the subject of justice), "What?" (the object of justice), and "How?" (equity standard(s) and concepts) was filled out for each study. @fig-A2 contains the template that was input into *Covidence* and used throughout.

@@

# The lay of the land: a summary of findings {#sect4}

@@

```{r plotting-year-cont-plot-fig3}
#| label: fig-fig3
#| fig-cap: "Papers included in the review by year of publication and case study continent."
year_cont_plot
ggsave("figures/year_cont_plot.png")
```

@@

```{r plotting-population-group-plot-fig4}
#| label: fig-fig4
#| fig-cap: "The proportion of papers that focus on each type of population group. Categories for population groups were generated upon data extraction."
pop_plot
ggsave("figures/pop_plot.png")
```

@@

```{r plotting-mode-plot-fig5, fig.width=5}
#| label: fig-fig5
#| fig-cap: "The proportion of papers that investiage each type of mode. Categories for modes were generated upon data extraction."
mode_plot
ggsave("figures/mode_plot.png")
```

@@

In terms of papers that study specific destinations, healthcare services (`r (dest_count$count[6]/sum(dest_count$count)) |> percent()`) and employment (`r (dest_count$count[5]/sum(dest_count$count)) |> percent()`) are activities commonly studied. Papers that exclusively focus on healthcare typically originate from the healthcare planning literature, and look to inform planners about disparities in services and corrective actions. For instance, @wangModelingSpatialAccessibility2011 model access of Chinese-language speakers in Toronto (Canada) to Mandarin-speaking family physicians. These authors infer an inequity from the spatial mismatch between language competent healthcare providers and Chinese speakers. Papers that exclusively focus on employment typically focus on these trips as they are the most common trip purpose and are often correlated with other trip activities like shops, recreation, and other services. For instance, @allenSizingTransportPoverty2019a propose a low employment-based accessibility threshold and a composite population vulnerability index to identify transportation-poor neighborhoods in urban Canada. Papers that focus on healthcare and employment typically source data from representative travel surveys/diaries, census data, and points-of-interest databases. In other words, these studies often benefit from well developed and institutional data that represents 'typical trips', especially in the Global North where these data are more readily available.

@@

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig6
#| fig-cap: "The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category."
standards_conc_percentages_plot_and_table
ggsave("figures/standards_conc_percentages_plot_and_table.png")
```
<!--`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards? what types of methods?", ".")`-->
We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice. Broadly speaking, some trends emerge in terms of the methods used depending on the way equity is conceptualized and the type of standard suggested; See @fig-fig6 for definitions of of conceptualizations along the horizontal axis and types of standards along the vertical axis. These categories emerged as we reviewed the literature and we use them to discuss trends; they are by no means an exhaustive list of equity conceptualizations and types of standards.

@@

# A critical appraisal of the lay of the land {#sect5}

@@

Most papers within our corpus (`r ((case_cont_count$count[4] + case_cont_count$count[5]+ case_cont_count$count[6])/sum(case_cont_count$count)) |> percent()`) focus on case studies in the Global North. Though their subject matter is varied, their spatial context mainly pertains to North American and Europe, and thus more often than not, in conversation with more developed and formal government transport planning apparatuses and technologies [e.g., planning for equitable high-speed rail, as in @monzonEfficiencySpatialEquity2013; autonomous vehicle technology, as in @eppenbergerOpportunitySharedAutonomous2021; or on the public consultation processes, as in @reddySafeguardingMinorityCivil2010]).

@@

Temporally, we see the literature from the Global North and Global South in our corupus as laying on different transport inequity continuums. Literature from the Global North is concerned with particularly disadvantaged groups and creating indicators that may be used to guide the remedy of formal processes that ensure access to societal benefits e.g., [@cuiSpatialAccessPublic2020]. Global North countries often surpass international sufficiency thresholds related to transport and public health (e.g., @carrierApplicationThreeMethods2014 measures NO_{2} road-side air pollution and find no levels are below the WHO guidelines) but advocate for the reduction in disproportional impact of air pollution on lower-income groups. These international guidelines, while perhaps being relevant at a time of lower transport-related development, are no longer relevant for some areas in the Global North. Along this line, inequities in the Global North are rising - high-income countries contain the wealthiest populations that have the most mobility (and produce associated carbon externalities) [@chancelCarbonInequalityKyoto2015].

@@

So what's missing from the "Who" question? We believe, more nuisances perspectives of inequities are needed. We must go beyond the focus on the low-income transit rider; positioning intersectionality of socio-economic and mode-use to create community-based definitions of equity and tailored standards are critical.

@@

# Moving forward: calls for action {#sect6}

@@

# Final remarks {#sect7}
We began this report with the history-making suborbital flight of Jeff Bezos and his fellow passengers, all of them very wealthy people. We wondered why coverage of the event, in particular Bezos declarations planetside, were not met with more critical questions: who are the subjects of the marvelous future that Bezos fancies his trip will contribute to? All of the human species or only those who can afford space tourism? What are the benefits of space tourism and how are they distributed? Who are the recipients of the impacts?

@@

# Acknowledgements
This report would not have been possible without the help and inspiration of countless others not let listed. In the early stages of the methodological design preparation of this report, the support from a team of literature review specialists at University of Toronto was invaluable; particularly the Research Services and Liaison Librarians. These experts provided guidance in cornering down the topic, identifying guidance material on literature synthesis and relevant databases, and in introducing us to the world of knowledge synthesis. Though our path diverged from a traditional scoping review, choosing to place and discuss the reviewed equity literature in connection to notions of justice, the protocol published on September 2022 (found [here](https://osf.io/rsb92)) was followed with only minor changes -- a reduction in the databases searched due to resource constraints.

Furthermore, thank you for the hours of support from undergraduate and graduate research assistants who poured through title and abstract screening and helped determine initial stages of full-text eligibility. Reviewing such a quantity of literature was new to many of us, we are deeply thankful for your help: Jan Domalaon, Dancel Gayle, Eryn Maloney, Mahdis Moghadasi, Anika Munir, and Subaita Refaaf.

@@

# Appendix {#sect8}
